### 2017-03-29 thesis/intro/intro.tex
pretty much ready to submit - tmp commit

However, the use of \gls{bmi} as an accurate measure of body fat composition and/or representation of the metabolic state of an individual remains controversial.
In fact, many studies have suggested that the use of other measurements, such as waist-to-height and waist-to-hip ratios, to classify the patients into different categories, as these measurements were better predictors of the diseases related to obesity than \gls{bmi} \citep{Dalton2003,Gelber2008,Lee2008}.
With that said, \gls{bmi} is still widely used due to the ease of data collection compared to other measurements.
As a result, most of the publicly available clinical data has height and weight data of the patients from which the \gls{bmi} can be calculated.
% With that said, \gls{bmi} is still widely used due to the ease of data collection compared to other measurements and, as a result, most of the publicly available clinical data have  height and weight data of the patients from which one can calculate the \gls{bmi} information.

@@

In addition to this, the worldwide prevalence of childhood obesity has been increasing steadily since 2000 \citep{WHO2014}.
The prevalence of overweight children in those under 5 was estimated to rise from 6.3\% in 2013 to 11\% by 2025 if the current trend continued \citep{WHO2014}.
% The current obesity status of the adult population is high already, but the fact that more of the younger population who are overweight and obese poses a much greater risk for the health of the population in the future.
The current prevalence of obesity in the younger population poses an even greater risk to the health of the population in the future than the current prevalence of obesity in the adult population.

In \gls{nz}, obesity is a serious problem as approximately one third (31.6\%) of adults in \gls{nz} were classified as obese in 2015/2016 \citep{Health2016}.
The Pacific and M\=aori populations were affected the most, where 67\% and 47\% of the adults, respectively, were obese \citep{Health2016}.
\gls{nz} not only has a high prevalence of obesity compared to many other countries of the world, but also has a unique health problem that affects Pacific and M\=aori populations more than the rest of the \gls{nz} population.

@@

Though there are many hormones that regulate satiety, only leptin and insulin will be covered here.
% Leptin gene was first discovered by \citet{Zhang1994} and the discovery of the leptin receptor gene was followed shortly after by \citet{Tartaglia1995}.
% Together, the identification of the leptin and leptin receptor genes was a major break through in the field of obesity research.
Leptin is a hormone that signals satiety and reduces food intake, and is secreted mainly by the white adipose tissue \citep{Moustafa2013,Zhang1994}.
The circulating concentration of leptin in the body is proportional to the body fat stores \citep{Barsh2002, Moustafa2013}.
Leptin signals through the \gls{jak}/\gls{stat} pathway via the long form of the leptin receptor that is present in high expression in the hypothalamus \citep{Ghilardi1996,Lee1996}.
Depending on the type of neurons the leptin receptor is expressed, leptin stimulate or inhibit these neurons and ultimately produces an anorexogenic signal to the downstream effector neurons \citep{Bell2005}.

@@

The exact role of insulin in the satiety control was not particularly clear even though there was evidence of increased food intake with neuron-specific deletion of insulin receptor \citep{Barsh2002,Bruning2000}.
The fact that the insulin level, like leptin, was proportional to the fat mass also suggested the role of insulin in satiety and body weight control \citep{Barsh2002, Bruning2000, Woods1979}.
This was perhaps due to the fact that leptin and insulin used two different signalling pathways in the cell: leptin used the \gls{jak}/\gls{stat} signalling pathway, whereas insulin used the \gls{pi3k} pathway for signalling \citep{Ghilardi1996}.
However, leptin, and subsequently insulin, was shown to activate ATP-dep\-endent K$^+$ channel in glucose-responsive neurons in the hypothalamus \citep{Spanswick1997, Spanswick2000}.
The activation of the K$^+$ channel provided evidence for the two pathways to act similarly on the cell, at least for an acute response, suggesting a convergence in the pathway response by the two hormones, and therefore their role in satiety control.

@@

An example of a mutation that causes severe early-onset obesity is a mutation in the leptin gene.
\citet{Montague1997} found a single nucleotide deletion mutation at position 133 of the leptin gene in children with early-onset obesity.
They further showed that the mutation caused a frameshift mutation in the protein, resulting in an inactive form of leptin that was not able to be secreted \citep{Montague1997}.
As a result, children with this mutation were constantly hungry due to the lack of anorexogenic signals and became obese at a very early stage of their lives.
% Though these sort of mutations are very rare and affects only those individuals that have these mutations, it had a profound impact in the field of obesity research and contributed significantly to the knowledge of the mechanism of the development of obesity.

Polygenic obesity is where a combination of variants in multiple genes or multiple regions of the genome (together with the effect from the environmental factors) causes obesity in the individual \citep{Moustafa2013}.
Unlike in monogenic obesity where a single mutation in a gene causes severe obesity, the variants in some of the genes or genomic regions involved in polygenic obesity have little effect on the obesity status of the individual by itself, but in combination with many other variants and together with environmental factors, it has a serious obesogenic effect on the individual.

There have been many studies that have investigated the association of genes and/or genomic regions with obesity.
Of those, many of the \glspl{gwas} published in 2007 showed strong evidence of the \gls{fto} gene with \gls{bmi} and obesity  \citep{Dina2007,Frayling2007,Gerken2007,Scuteri2007}.
Until these studies came out, there has been no common genes or genomic regions associated with obesity that was reproducible and reliable \citep{Frayling2007}.
Thus, these studies were the first evidence of common variants that contributed to the polygenic obesity phenotype in the population.

@@

It is clear that the individual-level factors cannot be accounted for the global epidemic of obesity that the world currently observes.
For this reason, obesity is thought to be caused mainly by the environmental factors that act on a population-level.
The first key factor that contributes to the global rise in the proportion of the population that is obese is the availability and abundance of certain foods in the country.
With the help from government trade policies, food and other goods have become easier to trade between countries, helping the economic growth of many countries around the world \citep{Kearney2010}.
This led to a greater abundance of food in general, a greater number of food choices and shifted the overall nutritional status of countries \citep{Malik2013}.
This shift in food choice and availability (and the economic benefit associated with the trades) had a positive impact on countries, but the growth in the food market also resulted in the population-wide rise in obesity.

As an example, in the \gls{usa}, the costs of corn and soy are low due to the fact that they are the raw ingredients for most processed food and beverages, such as high fructose corn syrups used to sweeten many soft drinks in the \gls{usa} \citep{Malik2013}.
Moreover, corn and soy are also the main food for the livestock, which results in lower prices for meat \citep{Malik2013}.
On the other hand, the prices of fruits and vegetables remain expensive due to the lack of support by the government to lower the cost associated with the production \citep{Malik2013}.
As a result, more people are inclined to consume foods that are cheap, have little nutritional value and high in energy, than foods that are expensive, nutritional and healthy alternatives.
Thus, the population is more likely to become obese by making poor food choice.

The second factor to consider is the behavioural changes led by urbanisation.
As more people take up urban lifestyle, they are exposed to new environments where there is a better range of food selection, as well as technological and mechanical advancement in transport and other daily chores that may not be available in rural areas.
Although there are many advantages involved with urbanisation, such as access to developed health care systems, education and advanced technologies, the lifestyle change also imposes negative health behaviour that ultimately can lead to positive energy balance and obesity \citep{Malik2013}.
Lack of physical activity is one of the obvious components linked with urbanisation.
In many urban cities, the recommended level of physical exercise is often difficult to achieve and maintain due to the shift towards sedentary behaviours that are encouraged by developed transport systems, automated household chores and indoor entertainment \citep{Malik2013}.

>>>>

However, the use of \gls{bmi} as an accurate measure of body fat composition and/or representation of the metabolic state of an individual remains controversial.
In fact, many studies have suggested that the use of other measurements, such as waist-to-height and waist-to-hip ratios, to classify the patients into different categories, as these measurements were better predictors of the diseases related to obesity than \gls{bmi} \citep{Dalton2003,Gelber2008,Lee2008}.
With that said, \gls{bmi} is still widely used due to the ease of data collection compared to other measurements.
As a result, most of the publicly available clinical data has height and weight data of the patients from which the \gls{bmi} can be calculated.

@@

In addition to this, the worldwide prevalence of childhood obesity has been increasing steadily since 2000 \citep{WHO2014}.
The prevalence of overweight children in those under 5 was estimated to rise from 6.3\% in 2013 to 11\% by 2025 if the current trend continued \citep{WHO2014}.
The current prevalence of obesity in the younger population poses an even greater risk to the health of the population in the future than the current prevalence of obesity in the adult population.

In \gls{nz}, obesity is a serious problem as approximately one third (31.6\%) of adults in \gls{nz} were classified as obese in 2015/2016 \citep{Health2016}.
The Pacific and M\=aori populations were affected the most, where 67\% and 47\% of the adults, respectively, were obese \citep{Health2016}.
\gls{nz} not only has a high prevalence of obesity compared to many other countries, but also has a unique health problem that affects Pacific and M\=aori populations disproportionately than the rest of the \gls{nz} population.

@@

Though there are many hormones that regulate satiety, only leptin and insulin will be covered here.
Leptin is a hormone that signals satiety and reduces food intake, and is secreted mainly by the white adipose tissue \citep{Moustafa2013,Zhang1994}.
The circulating concentration of leptin in the body is proportional to the body fat stores \citep{Barsh2002, Moustafa2013}.
Leptin signals through the \gls{jak}/\gls{stat} pathway via the long form of the leptin receptor that is present in high expression in the hypothalamus \citep{Ghilardi1996,Lee1996}.
Depending on the type of neurons the leptin receptor is expressed on, leptin stimulates or inhibits these neurons and ultimately produces an anorexogenic signal to the downstream effector neurons \citep{Bell2005}.

@@

The exact role of insulin in the satiety control was not particularly clear even though there was evidence of increased food intake with neuron-specific deletion of insulin receptor \citep{Barsh2002,Bruning2000}.
The fact that the insulin level, like leptin, was proportional to the fat mass also suggested the role of insulin in satiety and body weight control \citep{Barsh2002, Bruning2000, Woods1979}.
This was perhaps due to the fact that leptin and insulin used two different signalling pathways in the cell: leptin used the \gls{jak}/\gls{stat} signalling pathway, whereas insulin used the \gls{pi3k} pathway for signalling \citep{Ghilardi1996}.
However, leptin, and subsequently insulin, was shown to activate ATP-dep\-endent K$^+$ channel in glucose-responsive neurons in the hypothalamus \citep{Spanswick1997, Spanswick2000}.
The activation of the K$^+$ channel provided evidence for the two pathways to act similarly on the cell, at least for an acute response.
This suggests a convergence in the pathway response by the two hormones, and therefore their role in satiety control.

@@

An example of a mutation that causes severe early-onset obesity is a mutation in the leptin gene.
\citet{Montague1997} found a single nucleotide deletion mutation at position 133 of the leptin gene in children with early-onset obesity.
They further showed that the mutation caused a frameshift mutation in the protein, resulting in an inactive form of leptin that was not able to be secreted \citep{Montague1997}.
As a result, children with this mutation were constantly hungry due to the lack of anorexogenic signals and became obese at a very early stage of their lives.

Polygenic obesity is where a combination of variants in multiple genes or multiple regions of the genome (together with the effect from the environmental factors) causes obesity in the individual \citep{Moustafa2013}.
Unlike in monogenic obesity, where a single mutation in a gene causes severe obesity, the variants in some of the genes or genomic regions involved in polygenic obesity have little effect on the obesity status of the individual by itself.
However, in combination with many other variants and together with environmental factors, it has a serious obesogenic effect on the individual.

There have been many studies that have investigated the association of genes and/or genomic regions with obesity.
Of those, many of the \glspl{gwas} published in 2007 showed strong evidence of the \gls{fto} gene with \gls{bmi} and obesity  \citep{Dina2007,Frayling2007,Gerken2007,Scuteri2007}.
Until these studies came out, there has been no common genes or genomic regions found to be associated with obesity that was reproducible and reliable \citep{Frayling2007}.
Thus, these studies were the first evidence of common variants that contributed to the polygenic obesity phenotype in the population.

@@

It is clear that the individual-level factors cannot be accounted for the global epidemic of obesity that the world currently observes.
For this reason, obesity is thought to be caused mainly by the environmental factors that act on a population-level.
The first key factor that contributes to the global rise in the proportion of the population that is obese is the availability and abundance of certain foods in the country.
With the help from government free trade policies, food and other goods have become easier to trade between countries, helping the economic growth of many countries around the world \citep{Kearney2010}.
This has led to a greater abundance of food in general, a greater number of food choices and a change in the overall nutritional status of countries \citep{Malik2013}.
This shift in food choice and availability (and the economic benefit associated with the trade) had a positive impact on countries, but the growth in the food market also resulted in the population-wide rise in obesity.

As an example, in the \gls{usa}, the cost of corn and soy are low due to the fact that they are the raw ingredients for most processed food and beverages, such as high fructose corn syrups used to sweeten many soft drinks in the \gls{usa} \citep{Malik2013}.
Moreover, corn and soy are also the main food for livestock, which results in lower price for meat \citep{Malik2013}.
On the other hand, the prices of fruits and vegetables remain expensive due to the lack of support by the government to lower the cost associated with the production \citep{Malik2013}.
As a result, more people are inclined to consume food that are cheap, have little nutritional value and high in energy, than food that are expensive, nutritional and healthy alternatives.
Thus, the population is more likely to become obese by making poor food choice.

The second factor to consider is the behavioural changes led by urbanisation.
As more people take up urban lifestyle, they are exposed to new environments with a better range of food selection, as well as technological and mechanical advancement in transport and other daily chores that may not be available in rural areas.
Although urbanisation has many advantages, such as access to developed health care systems, education and advanced technologies, the lifestyle change also imposes negative health behaviour that ultimately can lead to positive energy balance and obesity \citep{Malik2013}.
Lack of physical activity is one of the obvious components linked with urbanisation.
In many urban cities, the recommended level of physical exercise is often difficult to achieve and maintain due to the shift towards sedentary behaviours that are encouraged by developed transport systems, automated household chores and indoor entertainment \citep{Malik2013}.

### 2017-03-29 thesis/misc/abstract.tex
pretty much ready to submit - tmp commit

It is now commonly understood that cancers are caused by dysregulation of various mechanisms or pathways that allow tumour cells to prolilferate, survive and migrate.
There have been many reports on the association between obesity and worsened cancer prognosis, suggesting that there may be distinct changes that underlie tumorigenesis and/or cancer progression that are specific to patients with higher \gls{bmi}.

In brief, this research aims to determine whether gene expression signatures exist that are specific to obesity across multiple cancer types, and to investigate whether there are any common pathways being dysregulated in cancers based on these genetic signatures.
Better understanding of the pathways being dysregulated in cancer cells in obese patients may lead to improved clinical decisions, and could thus contribute towards personalised treatment in the future.

>>>>

Obesity has been a major global problem for more than a decade, associated with many noncommunicable diseases such as  cancer.
The number of obese people, in both adults and children, has risen in every country of the world and the trend will likely to continue in the future.
Cancers are caused by dysregulation of various molecular pathways that allow tumour cells to proliferate, survive and migrate.
One of the difficulties associated with the treatment of cancers is the identification of the underlying biological pathways that drive tumorigenesis.
This research aims to determine whether gene expression signatures exist  that are specific to obesity across multiple cancer types, and to investigate whether there are any common pathways being dysregulated in cancers based on these genetic signatures.
There were no genetic signatures or genes differentially expressed between obese and non-obese patients that were common across multiple cancer types.
However, the Akt, \gls{egfr} and Src pathways may have a role in promoting the tumour progression in patients that are obese.
It is likely that there is some complex mechanism underlying the relationship between obesity and cancer.
Better understanding of the pathways being dysregulated in cancer cells in obese patients may lead to improved clinical decisions, and contribute towards personalised treatment in the future.
